Extracting The Power of Dependent Types

### 2021-09-27 paper.tex
Fix overfull lines
Pass for small optimizations for space and layout

@@

There are two common approaches to language embedding: deep and shallow.
Shallow embeddings define lan\-gua\-ges in terms of functions and operators
of the host language.  Programs in shallowly-embedded languages are programs
in the host language, so they are immediately run\-nable; types of the embedded
language are types of the host language, so the host typechecker is used for
checking embedded programs.  The downside of this approach is that we cannot
easily get the representation of embedded programs.

@@

%\paragraph{Embedding}
Defining deep embeddings with static guarantees are a common application
of dependent types~\cite{10.5555/647849.737066,CHAPMAN200921,
10.1007/978-3-540-74464-1_7,10.1145/3236785,10.1145/1863495.1863497}.
%
These embeddings usually also define semantics of the embedded
language and therefore allow us to reason about the correctness
of program transformations and optimisations.
%
While the fact that this is possible is impressive in theory, the resulting
encodings are often difficult to use in practice. In this paper
we instead aim for a more lightweight approach.
%
\citet{deepshallow} propose to solve this problem with
a combination of deep and shallow embeddings.  Their idea is to define
a small deep embedding and leverage type classes in Haskell to define
the rest of the language on top of that.  However, so far this idea has not yet
been applied to dependently typed embedded languages.

%\paragraph{Extraction}
The Coq proof assistant is equipped with extraction
capabilities~\cite{10.1007/978-3-540-69407-6_39,10.1007/3-540-39185-1_12},
which extracts functional code from Coq proofs (or programs).  The
default target language is OCaml, but a few other options were added
recently.
%
Likewise, Agda itself has a mechanism for defining custom backends, of
which the GHC backend is the most prominent.
%
Other proof assistants provide similar extraction tools as well.
%
The main difference from our approach in this paper
is that these extractors are written as plugins to the proof
assistant, while we implement our extractors directly in the proof
assistant itself.
%
While it would be possible to implement extractors presented in this
paper as Coq plugins or Agda backends, conceptually they are more
heavyweight.  Our extractors and programs can (in principle)
communicate with each other. In addition, as they are just Agda programs, they can
be reflected themselves and their structure can be leveraged.

%\paragraph{Dependently typed metaprogramming}
Several dependently-typed languages are equipped with metaprogramming
capabilities: Idris \cite{idris-refl}, Lean \cite{lean-refl},
Coq \cite{metacoq}, and Agda \cite{agda-refl}.  All of these
implement a similar API as described in this paper.  This
means that our proposed approach is immediately implementable
in these languages --- one simply needs to adjust extraction
to the particular API.
\citet{10.1145/3371071} introduce the Turnstile+ framework for
building dependently typed embedded DSLs and
shares the ideas advocated in this paper, suggesting that
our approach could work there as well.
\citet{10.1145/3371076} use MetaCoq to formally verify
the core type system of Coq. This combines nicely with
our approach, as we could use the verified core language
as a basis to verify our custom extractors.
\citet{10.1145/3372885.3373829} use MetaCoq to implement
a DSL combining deep and shallow approaches, in a way that
is quite similar to our own. While they are able to formally
reason about preservation of semantics (which we cannot do yet), it is unclear
whether their approach scales to dependently-typed embedded languages.

@@

In this paper we investigate the idea of developing
embedded programs hand-in-hand with custom
code generators for them. We solve the well-known
conundrum of choosing between deep and shallow
embedding by leveraging the power of
reflection.
This allows us to enjoy simplicity of writing programs
in shallow embedding.  Yet we keep full access to the internal
structure of the embedded programs, so that we could translate
them into the original language.

@@

The main advantage of our approach is twofold: first, it allows
you to implement all these things --- as well as the embedded programs
themselves --- \emph{side by side in a single language}, simplifying
the development. And secondly, it allows you to reuse the typechecker
and evaluator of the host language, saving you a lot of work.
While we showcase a single example, the approach has been applied
to several other languages: Kaleidoscope, a minimalist imperative language;
Single assignment C, a high-performance array language; and
APL (Array Programming Language), another array language with
heavily overloaded syntax.
For more details on these examples,
refer to~\citet{DBLP:journals/corr/abs-2105-10819}.

@@

\paragraph{Future work}
While our approach is flexible and powerful, it is
certainly not a magic bullet. A lot of things
need to be implemented for each target language: the embedding,
the encoding of the properties under consideration as types,
and the extractor.  For enforcing more complex properties, a number
of proofs needs to be written explicitly.

Right now, we cannot yet guarantee that the
extracted code preservers the semantics of the original
implementation. While we rarely see
fully-verified compiler backends in the real world,
our approach is very close to enabling this.
We would need a formal semantics of the reflected language
and the proof that reflected programs respect it.
While this is non-trivial, a system like Agda could do
this in principle.

>>>>

% Fixing overfull lines
\emergencystretch=10.0pt % default 0.0pt
\hyphenpenalty=500 % default 50
\exhyphenpenalty=500 % default 50

@@

There are two common approaches to language embedding: deep and shallow.
Shallow embeddings define lan\-gua\-ges in terms of functions and operators
of the host language.  Programs in shallowly embedded languages are programs
in the host language, so they are immediately run\-nable; types of the embedded
language are types of the host language, so the host typechecker is used for
checking embedded programs. However we cannot easily get the syntactic
representation of a shallowly embedded program.

@@

%\paragraph{Embedding}
Defining deep embeddings with static guarantees are a common application
of dependent types~\cite{10.5555/647849.737066,CHAPMAN200921,
10.1007/978-3-540-74464-1_7,10.1145/3236785,10.1145/1863495.1863497}.
%
These embeddings usually also define semantics of the embedded
language and therefore allow us to reason about the correctness
of program transformations and optimisations.
%
While the this is impressive in theory, the resulting
encodings are difficult to use in practice. In this paper
we instead aim for a more lightweight approach.
%
\citet{deepshallow} propose to solve this problem with
a combination of deep and shallow embeddings, by using
a small deep embedding and leveraging type classes in Haskell to define
the rest of the language on top of that.  However, so far this idea has not yet
been applied to dependently typed embedded languages.

%\paragraph{Extraction}
The Coq proof assistant is equipped with extraction
capabilities~\cite{10.1007/978-3-540-69407-6_39,10.1007/3-540-39185-1_12},
which extracts functional code from Coq proofs (or programs).  The
default target language is OCaml, but a few other options were added
recently.
%
Likewise, Agda itself has a mechanism for defining custom backends, of
which the GHC backend is the most prominent.
%
Other proof assistants provide similar extraction tools as well.
%
The main difference from our approach in this paper
is that these extractors are written as plugins to the proof
assistant, while we implement our extractors directly in the proof
assistant itself.
%
As a consequence, our extractors and programs can (in principle)
communicate with each other. In addition, as they are just Agda programs, they can
be reflected themselves and their structure can be leveraged.

%\paragraph{Dependently typed metaprogramming}
Several dependently-typed languages are equipped with metaprogramming
capabilities: Idris \cite{idris-refl}, Lean \cite{lean-refl},
Coq \cite{metacoq}, and Agda \cite{agda-refl}.  All of these
implement a similar API as described in this paper, and hence could be
used to implement our approach --- one simply needs to adjust
extraction to the particular API.
\citet{10.1145/3371071} introduce the Turnstile+ framework for
building dependently typed embedded DSLs and
shares the ideas advocated in this paper, suggesting that
our approach could work there as well.
\citet{10.1145/3371076} use MetaCoq to formally verify
the core type system of Coq. This combines nicely with
our approach, as we could use the verified core language
as a basis to verify our custom extractors.
\citet{10.1145/3372885.3373829} use MetaCoq to implement
a DSL combining deep and shallow approaches, in a way that
is quite similar to our own. While they are able to formally
reason about preservation of semantics (which we cannot do yet), it is unclear
whether their approach scales to dependently-typed embedded languages.

@@

In this paper we investigate the idea of developing
embedded programs hand-in-hand with custom
code generators for them. We solve the well-known
conundrum of choosing between deep and shallow
embedding by leveraging the power of
reflection.
This allows us to enjoy simplicity of writing programs
in shallow embedding, while keeping the ability to translate
them into the original language.

@@

The main advantage of our approach is twofold: first, it allows
you to implement all these things --- as well as the embedded programs
themselves --- \emph{side by side in a single language}, simplifying
the development. And secondly, it allows you to reuse the typechecker
and evaluator of the host language, saving you a lot of work.
While we showcase a single example, the approach has been applied
to several other languages: Kaleidoscope, a minimalist imperative language;
Single assignment C, a high-performance array language; and
APL (Array Programming Language), another array language with
heavily overloaded syntax~\cite{DBLP:journals/corr/abs-2105-10819}.

@@

\paragraph{Future work} While our approach is flexible, it is not a
magic bullet. Right now, we cannot yet guarantee that the extracted
code preservers the semantics of the original implementation. While we
rarely see fully-verified compiler backends in the real world, our
approach is very close to enabling this.  In future work, we would
like to give a formal semantics of the reflected language and the
proof that reflected programs respect it.  While this is non-trivial,
a system like Agda could do it in principle.

%\paragraph{Future work}
%While our approach is flexible and powerful, it is
%certainly not a magic bullet. A lot of things
%need to be implemented for each target language: the embedding,
%the encoding of the properties under consideration as types,
%and the extractor.  For enforcing more complex properties, a number
%of proofs needs to be written explicitly.
